\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Model

### 2021-01-15 Tex/accepted_preprint/accepted.tex
fixed typo in acknowledgements
removed comments

\title{\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Network Model\thanks{
		This work was supported in part by NSF grants SES-1558661, SES-1637089, SES-1619644, and CISE-1320219. We thank Rachel Hinkle, Benjamin Kassow, Lauren Santoro, and Michael Nelson for helpful feedback on this project. The replication code and data for this paper can be found at \citet{replicationdata}.}}
%\author{}
 % Toggle % below to blind and unblind the manuscript
%  \author[1]{Christian Schmid \thanks{cxs5700@psu.edu}}
%   \author[2]{Ted Hsuan Yun Chen\thanks{thc126@psu.edu}}
% \author[2]{Bruce A. Desmarais\thanks{bdesmarais@psu.edu}}
% \author[1]{David R. Hunter \thanks{dhunter@stat.psu.edu}}
% \affil[2]{Department of Political Science, Pennsylvania State University}
 %\affil[1]{Department of Statistics, Pennsylvania State University}

\author{%
  Christian S. Schmid\footnote{Department of Statistics, The Pennsylvania State University, schmid@psu.edu}%
  \and Ted Hsuan Yun Chen\footnote{Faculty of Social Sciences, University of Helsinki, ted.hsuanyun.chen@gmail.com}%
   \and Bruce A. Desmarais\footnote{Department of Political Science, The Pennsylvania State University, bdesmarais@psu.edu}%
%  \and David R. Hunter \footnote{Department of Statistics, The Pennsylvania State University, dhunter@stata.psu.edu}%
  }

@@

% \begin{figure}[bt]
% \centering
% \includegraphics[width = 0.85\textwidth,trim= 0cm 1cm 0cm 2cm,clip=true ]{citations_recip.pdf}
% \caption{Illustration of a reciprocal tie between two US Supreme Court opinions.}
% \label{fig:reciprocity}
% \end{figure}

@@

\section{The Citation Exponential Random Graph Model}
We develop a methodology that can be used to jointly test for the effects of covariates on citations---as have been studied in prior research, and test for dependence effects, as we hypothesize above. To accomplish this, we extend a model that has been developed to jointly represent covariate and dependence effects in network data, and has seen extensive application in recent political networks research---the exponential random graph model (ERGM) \citep[e.g.,][]{bratton2011networks, box2014evolution,duque2018recognizing,osei2018elite}.  %The ERGM, as it is currently implemented \citep{leifeld2017temporal,  tergm}, is designed to model longitudinal network data in which ties among a relatively stable set of nodes are modeled at multiple time points (e.g., the countries that are and are not defensively allied at time $t$, the legislators who have and have not cosponsored each others' legislation in the current session).  
The network structures for which ERGMs are currently designed are insufficient to account for the structure of citation networks.

@@

The likelihood function of the cERGM is given by
\begin{equation}
l(\bm{\theta},C_{\leq t}) =  \frac{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right] }{ \sum_{C_t^* \in \mathcal{C}_t} \exp \left[ {\bm{\theta}'\bm{h}(C^*_{t},C_{<t}) }\right]  },
\end{equation}
where $\bm{\theta}$ is a vector of real-valued parameters, and $\bm{h}(C_{t},C_{<t})$ is a vector of scalar-valued functions that each quantify a feature of the citation network (e.g., the relationship between citation ties and a case attribute, the number of mutual edges in the citation network). $\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right]$ is a positive weight that is proportional to the probability of observing any particular form of the citations and non-citations added to the network at time $t$.  The denominator of Equation 1 represents a normalizing constant, in which the positive weight is summed over all possible configurations of $C_{t}$, from the network in which the cases added to the network at time $t$ send no citations at all, to the network in which the cases added at time $t$ cite every possible case, and everything in between. %The probability of observing the citation network up to time $T$---the final time point in the dataset---is given by the product over the probability of adding the citations at each time point $t$, conditional on the citations that were added to the network previously.

@@

\begin{figure}[bt]
	\centering
	\begin{minipage}{.37\linewidth}
		\centering
		\includegraphics[width = 0.99\textwidth]{degree_distribution.pdf}
		\includegraphics[width = 0.8\textwidth]{data_mat.png}%
	\end{minipage}%
	\begin{minipage}{.63\linewidth}
		\includegraphics[width = \textwidth]{courts_layer.png}
		% \includegraphics[width = \textwidth,clip=true,trim=0.5cm 3.75cm 0.75cm 0.5cm ]{courts_layer_updated.pdf}
	\end{minipage}
	\caption{Supreme Court Citation Network, 1937-2015. Network visualization on the right. Nodes are Supreme Court cases, color-coded based on the chief justice presiding over the court. On the top left is the in- and outdegree distribution of the network. There are cases with an in- or outdegree  $>$50, but they are not captured in this figure. The bottom left shows the citation data in adjacency matrix format following \autoref{fig:ctergm}.}
	\label{degree_dist}
\end{figure}

% \begin{figure}[bt]
% \centering
% \includegraphics[scale = 0.45 ]{degree_distribution.pdf}
% \caption{The in- and outdegree distribution of the Supreme Court Citation Network from 1937 - 2015. There are cases with an in- or outdegree  $>$50, but they are not captured in this figure.}
%  \label{degree_dist}
% \end{figure}

% \begin{figure}[h]
% \centering
% \includegraphics[scale=0.20,clip=true,trim=.5cm 2cm 0cm 0cm ]{courts_layer_updated.pdf}
% \caption{ Supreme Court Citation Network, 1937-2015. Nodes are Supreme Court cases, color-coded based on the chief justice presiding over the court. Each network layer contains cases decided under a different chief justice. Because our data is temporally bound, cases for the Hughes and Roberts courts are not complete. }
%  \label{fig:networkviz}\end{figure}

@@

The first covariate we incorporate into the model accounts for the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following \citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would mean that cases cite those that are ideologically similar. %WE COULD USE THIS DATASET AS A SECOND APPLICATION

%We include three sets of dummy variables that account for the issue areas of cases. In the first set of dummy variables---sender intercepts---the variable $X^{ij}$ indicates the issue area of the sending case ($i$) (e.g., for the Criminal Procedure indicator, $X_{ij} = 1$ if case $i$ is a criminal procedure case, and $0$ otherwise). The second set of dummy variables represent receiver issue area intercepts in which $X^{ij}$ indicates the issue of the receiver case. The third set of dummy variables account for the rate of citation within each issue area and between each combination of issue areas.  There is a separate dummy variable for each directed pair of issue areas. For example the A$\rightarrow$B dummy variable $X_{ij} = 1$ if case $i$ is issue area A and case $j$ is issue area B, and 0 otherwise. Note that, given the high dimension of these dummy variables, we omit one each of the sender and recipient indicators, and 28 of the 196 mixing rate indicators. Unfortunately, this leaves these indicators largely uninterpretable, but these variables are nonetheless important to absorb variation in citations that is attributable to the legal domains of two cases in a dyad. The Issue area data comes from the Supreme Court Database (SCDB) \citep{spaeth2014supreme}. We include these variables because \citet{cross2010determinants} finds that the number of citations to Supreme Court opinions depends heavily on the issue area of the case. Table \ref{issue_area_coding} reports the issue areas covered in our data.

@@

%We also include a set of covariates that account for justice-specific variation in citations. We include two types of terms at the justice level. First, we include an indicator variable that models the rate at which justices cite themselves. This effect is modeled with a single indicator variable in which $X_{ij} = 1$ if the majority opinions in cases $i$ and $j$ were written by the same justice. Second, we use two sets of dummy variables to account for differences across justices both in terms of their tendencies to include citations in the opinions they write and to be cited by other opinions. The justice sender dummy variables indicate the identity of the justice who authored the majority or plurality opinion of case $i$. The justice recipient dummy variables indicate the identity of the justice who authored the majority or plurality opinion of case $j$. This set of variables is motivated by the use of citations by \citet{kosma1998measuring} to measure the influence of individual Supreme Court justices.

@@

%In order to account for the popularity effect, we add the \textit{geometrically weighted indegree distribution} (gwidegree) statistic to the model. The geometrically weighted distribution statistics were introduced by \citet{Hunter.2006} to counteract model degeneracy.

@@

%The results from the c-TERGM provide evidence that citations between US Supreme Court cases are driven both by exogenous case attributes and complex dependence processes. The scales of covariate and dependence effects are comparable. 
%The simple linear trends represent one limitation of the current analysis. In future work, more complex polynomials dynamical functions should be considered to determine whether effects truly flip signs.

@@

%We estimate the MCMLE for every term of the cERGM between 1950 and 2015. While the network for the 1950 term consists of $1962$ nodes, the size of the network monotonically increases to $10,020$ nodes for the 2015 term, making estimation more challenging as the network grows. The growth in the network also serves as our reason for estimating a separate model for each term. First, the network is large, with each term introducing the potential for hundreds-of-thousands of new citations. This means we have the statistical power necessary to estimate a separate set of parameter values for each term. Second, based on the findings of \citet{shalizi2013consistency}, we know that the set of parameter values that best fits a network should change as more nodes are added to the network. This means it would be inappropriate to assume a constant set of parameter values as more cases are added each term.

@@

\begin{figure}[bt]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{GOF_1950}\\
		%\caption{Goodness-of-fit diagnostic for the 1950 network.}
		% \label{GOF_1950}
		\vspace{1cm}
		\includegraphics[width=0.9\textwidth]{GOF_2015}
		\vspace{0.1cm}
		\caption{Goodness-of-fit diagnostic for the 1950 network (top) and the 2015 network (bottom).}
		\label{GOF}
		\vspace{-.25cm}
	\end{center}
\end{figure}

>>>>

\title{\Large Generative Dynamics of Supreme Court Citations: \\ Analysis with a New Statistical Network Model\thanks{
		This work was supported in part by NSF grants SES-1558661, SES-1637089, SES-1619644, and CISE-1320219. We thank Rachael Hinkle, Benjamin Kassow, Lauren Santoro, and Michael Nelson for helpful feedback on this project. The replication code and data for this paper can be found at \citet{replicationdata}.}}

\author{%
  Christian S. Schmid\footnote{Department of Statistics, The Pennsylvania State University, schmid@psu.edu}%
  \and Ted Hsuan Yun Chen\footnote{Faculty of Social Sciences, University of Helsinki, ted.hsuanyun.chen@gmail.com}%
   \and Bruce A. Desmarais\footnote{Department of Political Science, The Pennsylvania State University, bdesmarais@psu.edu}%
  }

@@

@@

\section{The Citation Exponential Random Graph Model}
We develop a methodology that can be used to jointly test for the effects of covariates on citations---as have been studied in prior research, and test for dependence effects, as we hypothesize above. To accomplish this, we extend a model that has been developed to jointly represent covariate and dependence effects in network data, and has seen extensive application in recent political networks research---the exponential random graph model (ERGM) \citep[e.g.,][]{bratton2011networks, box2014evolution,duque2018recognizing,osei2018elite}. The network structures for which ERGMs are currently designed are insufficient to account for the structure of citation networks.

@@

The likelihood function of the cERGM is given by
\begin{equation}
l(\bm{\theta},C_{\leq t}) =  \frac{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right] }{ \sum_{C_t^* \in \mathcal{C}_t} \exp \left[ {\bm{\theta}'\bm{h}(C^*_{t},C_{<t}) }\right]  },
\end{equation}
where $\bm{\theta}$ is a vector of real-valued parameters, and $\bm{h}(C_{t},C_{<t})$ is a vector of scalar-valued functions that each quantify a feature of the citation network (e.g., the relationship between citation ties and a case attribute, the number of mutual edges in the citation network). $\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}) } \right]$ is a positive weight that is proportional to the probability of observing any particular form of the citations and non-citations added to the network at time $t$.  The denominator of Equation 1 represents a normalizing constant, in which the positive weight is summed over all possible configurations of $C_{t}$, from the network in which the cases added to the network at time $t$ send no citations at all, to the network in which the cases added at time $t$ cite every possible case, and everything in between.

@@

\begin{figure}[bt]
	\centering
	\begin{minipage}{.37\linewidth}
		\centering
		\includegraphics[width = 0.99\textwidth]{degree_distribution.pdf}
		\includegraphics[width = 0.8\textwidth]{data_mat.png}%
	\end{minipage}%
	\begin{minipage}{.63\linewidth}
		\includegraphics[width = \textwidth]{courts_layer.png}
	\end{minipage}
	\caption{Supreme Court Citation Network, 1937-2015. Network visualization on the right. Nodes are Supreme Court cases, color-coded based on the chief justice presiding over the court. On the top left is the in- and outdegree distribution of the network. There are cases with an in- or outdegree  $>$50, but they are not captured in this figure. The bottom left shows the citation data in adjacency matrix format following \autoref{fig:ctergm}.}
	\label{degree_dist}
\end{figure}

@@

The first covariate we incorporate into the model accounts for the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following \citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would mean that cases cite those that are ideologically similar.

@@

@@

@@

@@

@@

\begin{figure}[bt]
	\begin{center}
		\includegraphics[width=0.9\textwidth]{GOF_1950}\\
		\vspace{1cm}
		\includegraphics[width=0.9\textwidth]{GOF_2015}
		\vspace{0.1cm}
		\caption{Goodness-of-fit diagnostic for the 1950 network (top) and the 2015 network (bottom).}
		\label{GOF}
		\vspace{-.25cm}
	\end{center}
\end{figure}
